---
title: "Changing Government Budget allocation during current Geopolitical stance"
subtitle: "Assignment 4 ETC5512"
author: "Aryaman Dhindsa - 35683422"
format: html
bibliography: references.bib
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data Details
### Question

How much does the world governments spending changed during the current geopolitical events in comparison to the last 2 decades?

*The world is currently divided into different ideologies such as during the mid-20th century and looking at budget allocation we can see where the future is headed as government spending shows the picture of future plans and preparations of countries*

### Data

The data provided my `WDI` is a great source as it is an aggregation of whole of the world government data that is available on economy, people, status and spending provided by the world Bank and is considered as the most reliable source. The API provided in R i.e. `package(WDI)` is a quick way to get all the needed variables without having to download the data.

#### Data Metadata

The analysis uses data obtained from the World Bank's World Development Indicators (WDI) database. Data are accessed programmatically through the `WDI` package in R, which provides an interface to the World Bank Data API.

| Attribute | Description |
|------------|------------|
| **Dataset** | World Development Indicators (WDI)[@wdicite] |
| **Source** | World Bank Group |
| **Access Method** | World Bank Data API accessed through the `WDI` package in R |
| **Geographic Coverage** | 260+ countries, territories, and regional aggregates |
| **Granularity** | National level |
| **Temporal Coverage** | 1960–2025 |
| **Frequency** | Annual |
| **Update Frequency** | Quarterly (April, July, September, December) |
| **Regularly Updated** | Yes |
| **Languages Available** | Arabic, English, French, Chinese, Spanish |
| **License** | Creative Commons Attribution 4.0 International (CC BY 4.0) |
| **First Published** | 1 April 2005 |
| **Last Updated** | 8 April 2026 |

#### Information

All the variables and other variable related data information is provided in `Data-dictionary.xlsx` present in the *data* folder.

All other information related to author, modified date and all other project related information including licence, limitations, data privacy is present in the `README.md` file.

#### Data 

Getting the package *Tidyverse*[@tidyverse] ready of R[@r] and loading the API for world development indicators[@worldbankapi].
```{r setup}
#| warning: false
#| error: false
#| echo: false
#| message: false

# Install packages if not already installed
if (!requireNamespace("WDI", quietly = TRUE)) install.packages("WDI")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")

library(WDI)
library(tidyverse)
library(maps)
library(viridis)
library(scales)
library(kableExtra)
library(maps)
```
Getting the required variables as needed for analysis:
```{r data}
#| cache: true

indicators <- c(
  mil_gdp    = "MS.MIL.XPND.GD.ZS",
  edu_gdp    = "SE.XPD.TOTL.GD.ZS",
  health_gdp = "SH.XPD.GHED.GD.ZS",
  gdp_usd    = "NY.GDP.MKTP.CD"
)

raw <- WDI(
  country   = "all",
  indicator = indicators,
  start     = 2005,
  end       = 2025,
  extra     = TRUE
)

countries <- raw |>
  filter(region != "Aggregates", !is.na(iso2c))

world_avg <- countries |>
  # drop rows where gdp_usd is NA so weighted.mean never gets an all-NA weight vector
  filter(!is.na(gdp_usd)) |>
  group_by(year) |>
  summarise(
    # wieghted mean is used to scale the budgets and spending means for rich and poor countries or big or small countries.
    Military   = weighted.mean(mil_gdp,    gdp_usd, na.rm = TRUE),
    Health     = weighted.mean(health_gdp, gdp_usd, na.rm = TRUE),
    Education  = weighted.mean(edu_gdp,    gdp_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = c(Military, Health, Education),
    names_to  = "sector",
    values_to = "pct_gdp"
  )

latest_mil <- countries |>
  filter(!is.na(mil_gdp)) |>
  group_by(country, iso2c) |>
  slice_max(year, n = 1) |>
  ungroup() |>
  select(country, iso2c, year, mil_gdp)
```
Sample of each dataset:
```{r}
raw |> head(5) |> kable()
world_avg |> head(5) |> kable()
latest_mil |> head(5) |> kable()
```

::: {#refs}
:::

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Why to Look at the Budgets?

Budget allocation is the most important aspect to examine if we want to understand the future plans of a country. Where governments choose to spend — on guns or schools, on missiles or medicines — is not merely an accounting decision. It is a statement of intent, a map of fears and ambitions, and a portrait of the world order that those in power are preparing for.

### What does main segments of budget tells about the current geopolitical conditions?

The last two decades have given us two fundamentally different eras of government spending. The 2000s and early 2010s were broadly a period of globalisation optimism — governments invested in healthcare systems, education, and social infrastructure. Then came the fractures: Russia's annexation of Crimea in 2014, a global pandemic in 2020, a full-scale land war in Europe in 2022, and a grinding conflict in the Middle East from 2023 onward. The budgets that followed tell a story of a world that no longer feels safe.

The other conclusion is that *The need to spend in Health and Military seems to be more important as for survival than investing in Education for future prosperity.*

- **Military expenditure** — as a share of GDP and in current USD
- **Government health expenditure** — as a share of GDP
- **Public spending on education** — as a share of GDP

Taken together, these three sectors reveal the essential tension in every national budget: how much to invest in security versus the people it is meant to protect.

```{r line-chart}
#| warning: false
#| message: false
#| echo: false
#| label: fig-spending-by-section
#| fig-cap: "GDP-weighted global average government spending by sector (% of GDP), 2000–2023. Source: World Bank WDI."

ggplot(world_avg, aes(x = year, y = pct_gdp, colour = sector)) +
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  geom_point(size = 1.8, na.rm = TRUE) +
  scale_colour_manual(
    values = c("Military" = "#C0392B", "Health" = "#2980B9", "Education" = "#27AE60")
  ) +
  labs(
    title   = "Military vs. Social Spending (2000–2023)",
    x       = NULL,
    y       = "% of GDP",
    colour  = "Sector",
    caption = "Source: World Bank WDI"
  ) +

  labs(
    title    = "The Great Reallocation: Military vs. Social Spending (2000–2023)",
    subtitle = "GDP-weighted world average — Military spending inflects upward after 2014, while social investment stagnates",
    x        = NULL,
    y        = "% of GDP",
    caption  = "Source: World Bank WDI"
  )+
  theme_minimal()
```
The @fig-spending-by-section gives us an overview that the investments in health have increased whereas the budget for education decreased by a lot. The important thing to notice here is the consistent increase in military spending after 2021. The war of Ukraine and Russia started in Feburay 2022[@wikipedia_russo_ukrainian_war_2026].

Now lets look as per the regions to get a more clear understanding of the situation:
```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-region-spending
#| fig-cap: "Government spending by region as % of GDP (2005–2025). Source: World Bank WDI."

region_avg <- countries |>
  filter(!is.na(gdp_usd)) |>
  group_by(region, year) |>
  summarise(
    Military  = weighted.mean(mil_gdp, gdp_usd, na.rm = TRUE),
    Health    = weighted.mean(health_gdp, gdp_usd, na.rm = TRUE),
    Education = weighted.mean(edu_gdp, gdp_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(c(Military, Health, Education), names_to = "sector", values_to = "pct_gdp")

ggplot(region_avg, aes(year, pct_gdp, colour = sector)) +
  geom_line(na.rm = TRUE) +
  facet_wrap(~region) +
  theme_minimal()
```
As shown in @fig-region-spending, military spending in the Middle East, North Africa, Afghanistan & Pakistan region rises sharply toward the end of the series, climbing from roughly 3-4% of GDP to nearly 7.5%, following an earlier surge around 2014-2015 to approximately 6% that likely coincided with regional conflicts in Syria and Yemen. The most recent spike plausibly reflects the escalation of conflict in the Middle East from 2023 onward.
By contrast, @fig-region-spending shows North America moving in the opposite direction on social spending: health expenditure falls from its 2020 pandemic-driven peak of around 10.5% of GDP to roughly 8%, while education spending declines even more steeply, from about 5% to under 2% in the final year. Military spending in North America also dips slightly at the end of the period.
Taken together, the figure suggests a reallocation of priorities — as the Middle East region increases military investment amid active conflict, North America appears to be drawing down both pandemic-era health spending and education funding. However, the sharp final-year drops in the North America panel may partly reflect provisional or incomplete data for the most recent year, so this should be treated as a preliminary observation pending confirmation with updated figures.

As estimated pre-analysis conclusion, it can be stated that mostly the changes can be seen in *Health* and *Defense*.So, lets look at them in detail fro the top 10 most powerful countries. Why? Because they are most influential countries in the world and their decisions have a big impact on other nations worldwide.

### Defense
**There are many wars going on around the world but only a few have an impact that can shake world order**
Budget increment by these countries can be a key indicator as they are already powerful enough to protect themselves from low or medium profile wars. But if they still increasing their budget, It can be a key insight as to know about the world order.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-top10-military
#| fig-cap: "Military expenditure (% of GDP) for the world's top 10 most powerful countries, 2021–2025. Source: World Bank WDI."


top10 <- c("United States", "China", "Russian Federation", "United Kingdom",
           "France", "Germany", "Japan", "India", "South Korea", "Israel")

top10_mil <- countries |>
  filter(country %in% top10, year > 2020, !is.na(mil_gdp))

ggplot(top10_mil, aes(year, mil_gdp, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "Military Spending (% of GDP) — Top 10 Powers, Post-2020",
       x = NULL, y = "% of GDP") +
  theme_minimal()
```
The @fig-top10-military shows that the 2 countries out 10 worlds most powerful countries increased their military spending in the last 5 years due to the wars they are in[@wikipedia_russo_ukrainian_war_2026]. Rest all are constant except Japan. Japan shows a slight increase in military budget due to its new policy of investing in its national weapon industries for defense[@reuters_japan_defence_2022].

### Health

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-top10-health
#| fig-cap: "Government health expenditure (% of GDP) for the world's top 10 most powerful countries, 2021–2025. Source: World Bank WDI."

top10_health <- countries |>
  filter(country %in% top10, year > 2020, !is.na(health_gdp))

ggplot(top10_health, aes(year, health_gdp, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "Health Spending (% of GDP) — Top 10 Powers, Post-2020",
       x = NULL, y = "% of GDP") +
  theme_minimal()
```
:::{.callout-note}
Data is not available post 2023 for some countries on health.
:::
The @fig-top10-health shows a general decline or flattening in health expenditure across most of the top 10 powers after their pandemic-era peaks, with the United States, Germany, and France easing back from elevated 2021 levels as emergency COVID-19 spending wound down. Russia and China remain at comparatively lower levels relative to their economic size, consistent with their prioritisation of military expenditure over healthcare investment during this period. Japan and South Korea, facing ageing populations, maintain relatively high and stable health spending despite the broader downward trend elsewhere.

### Conclusion

Looking at both the figures(@fig-top10-health and @fig-top10-military), military spending increasing primarily for countries directly engaged in conflict—Russia, and Middle Eastern nations—while remaining largely constant elsewhere, including most of Asia. Health spending, however, declined notably across European nations, Israel, and the USA, even as Asian countries like Japan and South Korea maintained stable levels. This divergence suggests that active warfare drives immediate defense reallocation, but its fiscal strain—particularly the wind-down of pandemic-era health investment—is felt more broadly across Western nations regardless of direct involvement, while Asian economies appear more insulated, maintaining steadier social spending priorities despite the shifting global security landscape

This pattern indicates that NATO countries are progressively reallocating budgets toward defense, while non-engaged nations elsewhere maintain steady spending priorities. It suggests Western countries are responding to instability in their neighbourhood by preparing for potential future conflicts, reflecting a broader sense of unrest across Europe and the Middle East.

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.

### AI acknowledgement

1. I used AI for research about the dataset variables.
2. Used AI for code improvement
3. Used AI for new insights.
4. Used AI for coding to focus on critical analysis and logical thinking.
Link of Claude: [here](https://claude.ai/share/1056a064-0a9b-4ffa-8c94-af537c88094d)
link of ChatGPT: [here](https://chatgpt.com/share/6a2cd6cf-e904-83ec-86dc-a082650d373e)

The link to the repository is [here](https://github.com/aryamandhindsa010/Open-Data-Analysis-World-spending-Analysis-WDI.git)
:::
