---
title: "Changing Government Budget allocation during current Geopolitical stance"
subtitle: "Assignment 4 ETC5512"
author: "Aryaman Dhindsa - 35683422"
format: html
bibliography: references.bib
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data Details
### Question

How much does the world governments spending changed during the current geopolitical events in comparison to the last 2 decades?

*The world is currently divided into different ideologies such as during the mid-20th century and looking at budget allocation we can see where the future is headed as government spending shows the picture of future plans and preparations of countries*

### Data

The data provided my `WDI` is a great source as it is an aggregation of whole of the world government data that is available on economy, people, status and spending provided by the world Bank and is considered as the most reliable source. The API provided in R i.e. `package(WDI)` is a quick way to get all the needed variables without having to download the data.

#### Data Metadata

The analysis uses data obtained from the World Bank's World Development Indicators (WDI) database. Data are accessed programmatically through the `WDI` package in R, which provides an interface to the World Bank Data API.

| Attribute | Description |
|------------|------------|
| **Dataset** | World Development Indicators (WDI)[@wdicite] |
| **Source** | World Bank Group |
| **Access Method** | World Bank Data API accessed through the `WDI` package in R |
| **Geographic Coverage** | 260+ countries, territories, and regional aggregates |
| **Granularity** | National level |
| **Temporal Coverage** | 1960–2025 |
| **Frequency** | Annual |
| **Update Frequency** | Quarterly (April, July, September, December) |
| **Regularly Updated** | Yes |
| **Languages Available** | Arabic, English, French, Chinese, Spanish |
| **License** | Creative Commons Attribution 4.0 International (CC BY 4.0) |
| **First Published** | 1 April 2005 |
| **Last Updated** | 8 April 2026 |

#### Information

All the variables and other variable related data information is provided in `Data-dictionary.xlsx` present in the *data* folder.

All other information related to author, modified date and all other project related information including licence, limitations, data privacy is present in the `README.md` file.

#### Data 

Getting the package *Tidyverse*[@tidyverse] ready of R[@r] and loading the API for world development indicators[@worldbankapi].
```{r setup}
#| warning: false
#| error: false
#| echo: false
#| message: false

# Install packages if not already installed
if (!requireNamespace("WDI", quietly = TRUE)) install.packages("WDI")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")

library(WDI)
library(tidyverse)
library(maps)
library(viridis)
library(scales)
library(kableExtra)
library(maps)
```
Getting the required variables as needed for analysis:
```{r data}
#| cache: true

indicators <- c(
  mil_gdp    = "MS.MIL.XPND.GD.ZS",
  edu_gdp    = "SE.XPD.TOTL.GD.ZS",
  health_gdp = "SH.XPD.GHED.GD.ZS",
  gdp_usd    = "NY.GDP.MKTP.CD"
)

raw <- WDI(
  country   = "all",
  indicator = indicators,
  start     = 2005,
  end       = 2025,
  extra     = TRUE
)

countries <- raw |>
  filter(region != "Aggregates", !is.na(iso2c))

world_avg <- countries |>
  # drop rows where gdp_usd is NA so weighted.mean never gets an all-NA weight vector
  filter(!is.na(gdp_usd)) |>
  group_by(year) |>
  summarise(
    # wieghted mean is used to scale the budgets and spending means for rich and poor countries or big or small countries.
    Military   = weighted.mean(mil_gdp,    gdp_usd, na.rm = TRUE),
    Health     = weighted.mean(health_gdp, gdp_usd, na.rm = TRUE),
    Education  = weighted.mean(edu_gdp,    gdp_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = c(Military, Health, Education),
    names_to  = "sector",
    values_to = "pct_gdp"
  )

latest_mil <- countries |>
  filter(!is.na(mil_gdp)) |>
  group_by(country, iso2c) |>
  slice_max(year, n = 1) |>
  ungroup() |>
  select(country, iso2c, year, mil_gdp)
```
Sample of each dataset:
```{r}
raw |> head(5) |> kable()
world_avg |> head(5) |> kable()
latest_mil |> head(5) |> kable()
```

::: {#refs}
:::

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Why to Look at the Budgets?

Budget allocation is the most important aspect to examine if we want to understand the future plans of a country. Where governments choose to spend — on guns or schools, on missiles or medicines — is not merely an accounting decision. It is a statement of intent, a map of fears and ambitions, and a portrait of the world order that those in power are preparing for.

### What does main segments of budget tells about the current geopolitical conditions?

The last two decades have given us two fundamentally different eras of government spending. The 2000s and early 2010s were broadly a period of globalisation optimism — governments invested in healthcare systems, education, and social infrastructure. Then came the fractures: Russia's annexation of Crimea in 2014, a global pandemic in 2020, a full-scale land war in Europe in 2022, and a grinding conflict in the Middle East from 2023 onward. The budgets that followed tell a story of a world that no longer feels safe.

The other conclusion is that *The need to spend in Health and Military seems to be more important as for survival than investing in Education for future prosperity.*

- **Military expenditure** — as a share of GDP and in current USD
- **Government health expenditure** — as a share of GDP
- **Public spending on education** — as a share of GDP

Taken together, these three sectors reveal the essential tension in every national budget: how much to invest in security versus the people it is meant to protect.

```{r line-chart}
#| warning: false
#| message: false
#| echo: false
#| label: fig-spending-by-section
#| fig-cap: "GDP-weighted global average government spending by sector (% of GDP), 2000–2023. Source: World Bank WDI."

ggplot(world_avg, aes(x = year, y = pct_gdp, colour = sector)) +
  geom_line(linewidth = 1.2, na.rm = TRUE) +
  geom_point(size = 1.8, na.rm = TRUE) +
  scale_colour_manual(
    values = c("Military" = "#C0392B", "Health" = "#2980B9", "Education" = "#27AE60")
  ) +
  labs(
    title   = "Military vs. Social Spending (2000–2023)",
    x       = NULL,
    y       = "% of GDP",
    colour  = "Sector",
    caption = "Source: World Bank WDI"
  ) +

  labs(
    title    = "The Great Reallocation: Military vs. Social Spending (2000–2023)",
    subtitle = "GDP-weighted world average — Military spending inflects upward after 2014, while social investment stagnates",
    x        = NULL,
    y        = "% of GDP",
    caption  = "Source: World Bank WDI"
  )+
  theme_minimal()
```
The @fig-spending-by-section gives us an overview that the investments in health have increased whereas the budget for education decreased by a lot. The important thing to notice here is the consistent increase in military spending after 2021. The war of Ukraine and Russia started in Feburay 2022[@wikipedia_russo_ukrainian_war_2026].

Now lets look as per the regions to get a more clear understanding of the situation:
```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-region-spending
#| fig-cap: "Government spending by region as % of GDP (2005–2025). Source: World Bank WDI."

region_avg <- countries |>
  filter(!is.na(gdp_usd)) |>
  group_by(region, year) |>
  summarise(
    Military  = weighted.mean(mil_gdp, gdp_usd, na.rm = TRUE),
    Health    = weighted.mean(health_gdp, gdp_usd, na.rm = TRUE),
    Education = weighted.mean(edu_gdp, gdp_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(c(Military, Health, Education), names_to = "sector", values_to = "pct_gdp")

ggplot(region_avg, aes(year, pct_gdp, colour = sector)) +
  geom_line(na.rm = TRUE) +
  facet_wrap(~region) +
  theme_minimal()
```
As shown in @fig-region-spending, military spending in the Middle East, North Africa, Afghanistan & Pakistan region rises sharply at the end of the series, following an earlier surge around 2014-2015 likely tied to conflicts in Syria and Yemen, with the most recent spike plausibly reflecting escalation since 2023. By contrast, North America moves in the opposite direction: health spending falls from its 2020 pandemic-driven peak, while education declines even more sharply. This suggests a reallocation toward security in conflict regions, though North America's final-year figures may be provisional and warrant confirmation with updated data.

As estimated pre-analysis conclusion, it can be stated that mostly the changes can be seen in *Health* and *Defense*.So, lets look at them in detail fro the top 10 most powerful countries. Why? Because they are most influential countries in the world and their decisions have a big impact on other nations worldwide.

### Defense
**There are many wars going on around the world but only a few have an impact that can shake world order**
Budget increment by these countries can be a key indicator as they are already powerful enough to protect themselves from low or medium profile wars. But if they still increasing their budget, It can be a key insight as to know about the world order.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-top10-military
#| fig-cap: "Military expenditure (% of GDP) for the world's top 10 most powerful countries, 2021–2025. Source: World Bank WDI."


top10 <- c("United States", "China", "Russian Federation", "United Kingdom",
           "France", "Germany", "Japan", "India", "South Korea", "Israel")

top10_mil <- countries |>
  filter(country %in% top10, year > 2005, year < 2023, !is.na(mil_gdp))

ggplot(top10_mil, aes(year, mil_gdp, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "Military Spending (% of GDP) — Top 10 Powers, Post-2020",
       x = NULL, y = "% of GDP") +
  theme_minimal()
```
As shown in @fig-top10-military, defense spending as a percentage of GDP has shifted notably for a few key nations over the last five years. Driven by active conflicts, the Russian Federation shows a sharp spike starting in 2021, while Israel—which consistently holds the highest baseline on the chart—also ticks upward at the end of the period. Conversely, the United States shows a gradual decline, while Japan maintains a very low but slightly increasing trajectory due to its updated defense policies. Spending for the remaining nations, including China, France, Germany, India, and the United Kingdom, has remained largely constant.

### Health

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| label: fig-top10-health
#| fig-cap: "Government health expenditure (% of GDP) for the world's top 10 most powerful countries, 2021–2025. Source: World Bank WDI."

top10_health <- countries |>
  filter(country %in% top10, year > 2005, year < 2023, !is.na(health_gdp))

ggplot(top10_health, aes(year, health_gdp, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "Health Spending (% of GDP) — Top 10 Powers, Post-2020",
       x = NULL, y = "% of GDP") +
  theme_minimal()
```
:::{.callout-note}
Data is not available post 2023 for some countries on health.
:::
The @fig-top10-health shows the United States, United Kingdom, Germany, and France clustering at elevated levels, peaking sharply toward the end of the series before the US and UK ease back, while Germany and France remain comparatively stable. Japan shows a steady upward climb throughout, ending as the highest spender. Russia and Israel show a sharp late spike, with Israel falling back afterward. China and India remain consistently low relative to the other powers, with China showing gradual growth and India staying nearly flat — both well below their economic and military significance, consistent with their relative prioritisation of defense over healthcare investment.

### Conclusion

Looking at both the figures(@fig-top10-health and @fig-top10-military), military spending increasing primarily for countries directly engaged in conflict—Russia, and Middle Eastern nations—while remaining largely constant elsewhere, including most of Asia. Health spending, however, declined notably across European nations, Israel, and the USA, even as Asian countries like Japan and South Korea maintained stable levels. This divergence suggests that active warfare drives immediate defense reallocation, but its fiscal strain—particularly the wind-down of pandemic-era health investment—is felt more broadly across Western nations regardless of direct involvement, while Asian economies appear more insulated, maintaining steadier social spending priorities despite the shifting global security landscape

This pattern indicates that NATO countries are progressively reallocating budgets toward defense, while non-engaged nations elsewhere maintain steady spending priorities. It suggests Western countries are responding to instability in their neighbourhood by preparing for potential future conflicts, reflecting a broader sense of unrest across Europe and the Middle East.

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### Question-5 

*Tell us about parts of your data processing or analysis that weren’t “sexy” and wouldn’t typically be included in a blog post. (e.g. Was their any data drudgery or time intensive wrangling? Were there any repetitive tasks or manual tasks? If it was easy, describe what made it easy?)*

Merging datasets from different sources is a notorious headache in data science, especially when it comes to country nomenclature. A primary hurdle in generating world data visualizations like @fig-top10-military involves reconciling disparate naming conventions between R's maps package and the World Bank’s World Development Indicators (WDI).

For instance, a country seamlessly recognized as "Bahamas" in the mapping geometry is designated as "Bahamas, The" within the WDI repository. Because these exact string matches fail behind the scenes, the corresponding spatial data often drops entirely from the final visualization. Due to the sheer volume of these inconsistencies and the tedious, manual string cleaning required to bridge the two packages, standardizing every mismatched nation was bypassed in this iteration. Consequently, the analysis focuses strictly on the major global powers cleanly supported by both data frameworks, ensuring a reliable visualization of military spending trends without letting minor syntax discrepancies stall the core pipeline.

### Question-6

*Were there any challenges that you faced in conducting this analysis. These may take the form of data limitations or coding challenges? (e.g. Was there anything in your analysis that you were not anticipating when you started? Did you have to change your intended scope? Did you need to master a new skill? Were there any problems you were proud of solving?)*

A significant data limitation emerged regarding temporal availability: while military expenditure data extended reliably into recent years, comprehensive global metrics for health and education spending were unavailable after 2023. This unexpected data lag made it impossible to conduct a synchronized, up-to-date comparative analysis across all three sectors for the final years of the timeline. Consequently, the scope was refined to prioritize the most current defense trends for major global powers where data was robust, highlighting how geopolitical events—rather than a full suite of social indicators—shaped recent budgetary shifts.

### Question-7

*Tell us about any imperfect parts of your work and how you would like to expand or improve this analysis in future? Be clear about any limitations or aspects of your analysis that fell beyond scope.*

While the current visualization effectively captures defense spending trends for major global powers, it remains imperfect due to data integration limits that restricted its scope. In the future, I want to expand this analysis to focus on specific high-conflict regions—particularly the immediate neighbors of Russia and Israel—to evaluate if and how surrounding geopolitical instability impacts regional military budgets.

The primary limitation preventing this deeper dive was the syntax mismatch between R's maps package and the World Development Indicators (WDI) data. Because smaller neighboring countries often suffered from mismatched nomenclature (e.g., failing to join properly during data merging), manually resolving every string discrepancy fell beyond the immediate scope of this iteration.

To improve this work, a crucial next step will be implementing a robust data-cleaning pipeline—utilizing tools like the countrycode package in R—to standardize all country names into ISO codes. Resolving these coding hurdles will allow the analysis to transition from a broad, macro-level overview of global superpowers to a precise, localized study on how regional conflicts alter the economic and defensive priorities of bordering nations.

### Question-8

*Tell us about how you used AI in your approach to your assignment. (e.g. Was AI consistently reliable? Did you provide AI with additional context? Was there anywhere where you questioned or verified the AI outputs? Was there anywhere we AI helped you learn something new?)*

Throughout this assignment, AI served as an iterative collaborator, primarily assisting with technical execution, dataset research, and code optimization. By leveraging AI to write, debug, and streamline complex R code, I was able to automate repetitive programming tasks. This significantly shifted my focus away from syntax troubleshooting and allowed me to dedicate more time to critical analysis, logical structuring, and interpreting the underlying geopolitical trends.

However, the AI was not blindly trusted. While it was highly reliable for optimizing code blocks, I consistently provided it with additional context—such as specific dataset schemas—and actively verified its outputs against the raw World Development Indicators (WDI) data to ensure accuracy. Beyond just code generation, the AI helped me learn new methodologies, offering fresh insights into how to structure a comparative analysis and suggest data-cleaning workflows. By treating AI as a tool for efficiency rather than a substitute for critical thought, I was able to deepen my understanding of the data while maintaining strict analytical accuracy.

Link of Claude: [here](https://claude.ai/share/1056a064-0a9b-4ffa-8c94-af537c88094d)
link of ChatGPT: [here](https://chatgpt.com/share/6a2cd6cf-e904-83ec-86dc-a082650d373e)

The link to the repository is [here](https://github.com/aryamandhindsa010/Open-Data-Analysis-World-spending-Analysis-WDI.git)
:::
